---
title: "CodeBook"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Code Description

First stage of this project was to install and load the necessary packages:  
`dplyr`  
`readR`  
`stringR`  

```{r install_packages,eval=FALSE}

install.packages("tidyverse")
install.packages("stringr")
library(tidyverse)
library(stringr)

```

It was necessary to download, save and unzip the files in a data directory:

```{r download_files, eval = FALSE}

if(!file.exists("./data")){
        dir.create("./data")
        }

if(!file.exists("./data/UCI HAR Dataset")){
        fileUrl <- "https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip"
        destfile <- "./data/dataset.zip"
        download.file(fileUrl, destfile = destfile, method = "curl" )
        unzip(zipfile = destfile, exdir = "./data")
        }
```

All the necessary files were loaded into R using `readR` package. Each experiment information was separated in 6 files divided in 2 sets (test and train) - two files with the activities information (`y_test` and `y_train`), two files with the measured data (`x_test` and `x_train`) and two files with the subject identification (`subject_test` and `subject_train`). I've also loaded a third file with the variables labels `features`, that was later used to name the measured variables:

```{r load_files, eval = FALSE}

x_test <- read_table("./data/UCI HAR Dataset/test/X_test.txt", col_names = FALSE)

x_train <- read_table("./data/UCI HAR Dataset/train/X_train.txt", col_names = FALSE)

y_test <- read_table("./data/UCI HAR Dataset/test/y_test.txt", col_names = "Code", col_types = cols(.default = col_factor(Code = col_factor(levels = c("1","2","3","4","5","6" )))))

y_train <- read_table("./data/UCI HAR Dataset/train/y_train.txt", col_names = "Code",  col_types = cols(.default = col_factor(Code = col_factor(levels = c("1","2","3","4","5","6" )))))

features <- read_table("./data/UCI HAR Dataset/features.txt", col_names = FALSE)

subject_test <- read_table("./data/UCI HAR Dataset/test/subject_test.txt", col_names = "Subject", col_types = cols(.default = col_number()))

subject_train <- read_table("./data/UCI HAR Dataset/train/subject_train.txt", col_names = "Subject", col_types = cols(.default = col_number()))

activities_labels <- read_table("./data/UCI HAR Dataset/activity_labels.txt", col_names = c("Code","Activity"), col_types = cols(Code = col_factor(levels = c("1","2","3","4","5","6" ))))
```

I've serarate each row in the feature dataset to extract only the variable name without the number preecededing it. 

```{r, eval = FALSE}

features <- separate(features,X1, sep = " ", into = c("Rank","Feature"))

colnames(x_test) <- features$Feature 

colnames(x_train) <- features$Feature

```

With all the files loaded and prepared, the next step was to bind the files together in 2 datasets. One with the `train` data and the other with the `test`data. 

```{r, eval = FALSE}

test_full <- bind_cols(subject_test,y_test,x_test) %>% mutate(Code = factor(Code, levels = c("1","2","3","4","5","6" )))

train_full <- bind_cols(subject_train,y_train,x_train) %>% mutate(Code = factor(Code, levels = c("1","2","3","4","5","6" )))
```

I've recoded the activity collumn with descriptive data instead of numbers

```{r, eval = FALSE}

test_full <- left_join(test_full, activities_labels, by = "Code" )

train_full <- left_join(train_full, activities_labels, by = "Code" )

```

I've created an index (`pattern`) to filter the variables that registered means and standard deviation acoording to instructions. I decided only to consider the variables which std() and mean() were found at the end of the variable label. So I created a regular expression ("((M|m)ean|(S|s)td)(\\(\\)-(X|Y|Z))$") to extract this data through `stringR`:str_subset command. Then train and test files were bind together and I renamed all the variables with descriptive (a kind long) labels.

```{r, eval = FALSE}

pattern <- "((M|m)ean|(S|s)td)(\\(\\)-(X|Y|Z))$"

index_test <- str_subset(names(test_full), pattern = pattern)

index_train <-  str_subset(names(train_full), pattern = pattern)

test <- test_full %>%
        select(Subject, Activity, all_of(index_test))

train <- train_full %>%
        select(Subject, Activity, all_of(index_train))

joined_table <- test %>%
        bind_rows(train) %>%
        arrange(Subject) %>% 
        mutate(Subject = Activity = parse_factor(Activity))

names(joined_table) <- str_replace(names(joined_table), "^t", "Time")

names(joined_table) <- str_replace(names(joined_table), "^f", "Frequency")

names(joined_table) <- str_replace(names(joined_table), "Acc", "Accelerometer")

names(joined_table) <- str_replace(names(joined_table), "Gyro", "Gyroscope")

names(joined_table) <- str_replace(names(joined_table), "Mag", "Magnitude")

names(joined_table) <- str_replace(names(joined_table), "BodyBody", "Body")

```

I've created a file, according to the project instructions with the tidy data:

```{r, eval = FALSE}
write.table(joined_table, file = "./data/tidy.txt", quote = FALSE, row.names = FALSE)
```

And finally, I've created a second, independent tidy data set with the average of each variable for each activity and each subject

```{r, eval = FALSE}
tidy_means <- joined_table %>%
        group_by(Subject,Activity) %>% summarise_all(.funs = mean)
```

# 2. Variables description

Variable                                      | Description              
----------------------------------------------|------------------------------------------------
Subject                                       | Each participan numbered from 1 to 30 [number]
Activity                                      | Activity during experiment (standing, walking, etc.) [factor]
TimeBodyAccelerometer                         | Mean and Standad deviation in X,Y,Z axis [number]
TimeGravityAccelerometer                      | Mean and Standad deviation in X,Y,Z axis [number]
TimeBodyAccelerometerJerk                     | Mean and Standad deviation in X,Y,Z axis [number]
TimeBodyGyroscope                             | Mean and Standad deviation in X,Y,Z axis [number]
TimeBodyGyroscopeJerk                         | Mean and Standad deviation in X,Y,Z axis [number]
FrequencyBodyAccelerometer                    | Mean and Standad deviation in X,Y,Z axis [number]
FrequencyBodyAccelerometerJerk                | Mean and Standad deviation in X,Y,Z axis [number]
FrequencyBodyGyroscope                        | Mean and Standad deviation in X,Y,Z axis [number]
----------------------------------------------------------------------------------------------

------------------------